---
title: "Introduction to R workshop"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# 1. Getting Started

R is a programming language -- in this workshop we'll use RStudio as an interface to explore data, write some code, and visualize results!

You can also use the cloud version for free - https://posit.cloud/plans/free. 

## Install R and RStudio

First, download both R (cran.r-project.org) and RStudio (posit.co/downloads/). 
Go to https://rstudio-education.github.io/hopr/starting.html for more information. 

## This document

Within the code chunks, complete the lines of code requested. When you're ready to evaluate your code, set each chunk to evaluate (`eval = TRUE`) -- 'knit' the document to create a nice pdf output for the entire document, or you can run specific code chunks by pressing the green triangle on the upper right hand side of that code chunk. 

# 2. The Basics

## Running code: console vs. script file

Navigate to the console in the bottom left of RStudio. Next to the `>` symbol, type `2*8` and press enter or return. 

Now try running:

```{r, eval = FALSE, echo = TRUE}

5 + "hello"

```

Your running history and the output for the code you run will show up in the console. 

To save lines of code (or run multiple lines), use a script file! 

## Creating variables

We use the `<-` operator to assign values to variables in R. 

### 1. Create a variable `my_age` in the code chunk below
```{r, echo = TRUE, eval= FALSE}

my_age <- 

```

### 2. Create a numeric vector `my_numbers` with three numbers separated by commas
```{r, echo = TRUE, eval = FALSE}

my_numbers <- c(_, _, _)

```

### 3. Create a variable that is equal to a word using ""

```{r, echo=TRUE, eval=FALSE}



```

### 4. Update my_age by adding 5 years

```{r, echo=TRUE, eval=FALSE}


```

## Variables and data classes

Variables can be a lot of different things in R! Unlike some other languages, there is no need to initialize variables before you assign values. To see what R is classifying a specific variable (or structure of data) as, use the `class()` function. 

### 1. Identify the data classes of the variables you just created

```{r, eval=FALSE, echo=TRUE}

class(variable1)
class(variable2)
class(variable3)

```

## Vectors

Earlier, we created a variable called `my_numbers` that contains multiple numbers -- this is called a "vector" in R, which is a structure of data (rather than a data class itself). 

We used the `c()` function to combine different numbers into this vector structure. 

### 1. Combine different words into a vector
```{r, echo=TRUE, eval=FALSE}

vector_of_words <- c()

```

## Summary statistics 

Now that we have a handful of values (in our numeric vector), we can do some summary statistics! We'll use a few base R functions: 

### 1. Find the `mean()`
```{r, echo=TRUE, eval=FALSE}


```

### 2. Find the `sum()`
```{r, echo=TRUE, eval=FALSE}


```

### 3. Find the `length()`
```{r, echo=TRUE, eval=FALSE}


```

## Indexing and logical filtering

There are often cases where we want to select a specific section of our data. In R, counting positions of a data structure is called indexing and choosing a specific piece of the data is finding a subsection or filtering the dataset. 

### 1. Find the second element of your `my_numbers` vector using []
```{r, echo=TRUE, eval=FALSE}


```

### 2. Pull out the 6th and 7th elements of this vector using one line of code
```{r, echo=TRUE, eval=FALSE}

example_vector <- sample(1:100, size = 10, replace = TRUE)

```

### 3. Pull out the values of `example_vector` that are greater than or equal to 10
```{r, echo=TRUE, eval=FALSE}


```

### 4. Save the result of #3 in a new variable
```{r, echo=TRUE, eval=FALSE}


```

# 3. Dataset example

Take a look at the documentation on the `beaver2` dataset in base R - note what the columns are named, the variables included, and the overall design of the study.

```{r, eval = FALSE, echo=TRUE}
?beaver2
```

## Exploring an example dataset 

You can use the `head()` function to see the first 6 rows of the dataset. Each row in the dataset is a measurement of body temperature (the entire dataset is one individual beaver), and the columns correspond to the time (`day` and `time`) of data collection, the response (`temp`), and an additional potentially explanatory variable (`activ` - values are 0 (no activity outside retreat) or 1 (activity outside retreat)). 

```{r, eval=TRUE, echo=TRUE}
head(beaver2)
```

Another few great exploratory functions for your dataset:
```{r, eval=TRUE, echo=TRUE}
str(beaver2) #"structure"
summary(beaver2)
```

What is the average body temperature? 
What is the range of body temperatures?

## Subset the dataframe {.unnumbered}

We know that the `activ` column indicates whether there was activity outside of a retreat, so perhaps we want to separate temperature measurements based on this activity. Figuring out how to subset our dataframe is useful here.

The `$` operator can be used to specify the name of a column in a dataframe when subsetting data.

### 1. Create a new variable that is our dataset of rows only where the beaver had activity outside the retreat 

Don't forget the comma in your square brackets!

```{r, eval=FALSE, echo=TRUE}

beaver_activity <- beaver2[] 

```

### 2. Create a new variable that is our dataset of rows only where the beaver had a body temperature greater than 37.5 C

```{r, eval=FALSE, echo=TRUE}

beaver_warm <- 

```

### 3. Combine the two conditions mentioned above for a new variable called `active_warm`
```{r, eval=FALSE, echo=TRUE}

```

## Base R visualizations

Visualizing our data can help us understand the structure even further. 

### 1. Histogram of body temperatures
```{r}
hist(beaver2$temp) 
```

Now we'll use the `boxplot()` function to look at the spread of body temperatures (`temp`) by detected activity (`activ`). 

### 2. Replace the `y_value` and `x_value` elements with the appropriate column names
```{r}
boxplot(y_value ~ x_value, data = beaver2,
        xlab = "Activity (0 = none detected, 1 = detected)",
        ylab = "Temperature",
        main = "This is a title")
```

### 3. At what time does the beaver reach its highest temperature? Why is the code below wrong?
```{r, echo=TRUE}

beaver2[max(beaver2$temp), ] #not correct!
max(beaver2$temp) #returns maximum temperature value

```

Try again with `which.max()`.

## More visualization

Here, we can use the graphing function included in the documentation to look at body temperature (`temp`) over the time of the study (progressive hours calculated using `day` and `time`). This function also includes a line showing presumably the high end of normal body temp (the `abline` line) and circles points where the beaver is noted as active (red circles, `points` line -- `col = 2` corresponds to red).

```{r, echo=TRUE}

beaver.plot <- function(bdat, ...) {
  nam <- deparse(substitute(bdat))
  with(bdat, {
    # Hours since start of day:
    hours <- time %/% 100 + 24*(day - day[1]) + (time %% 100)/60
    plot (hours, temp, type = "l", ...,
          main = paste(nam, "body temperature"))
    abline(h = 37.5, col = "gray", lty = 2)
    is.act <- activ == 1
    points(hours[is.act], temp[is.act], col = 2, cex = .8)
  })
}

beaver.plot(beaver2) #calling the function we just wrote

```

What trends of body temperature do you notice based on activity?

# 4. Writing functions

Functions are useful to write when we've copied and pasted code a few times!

### 1. Add a default value to the following function:
```{r, echo=TRUE, eval=FALSE}

celsius_to_kelvin <- function(temp_C) {
  temp_K <- temp_C + 273.15
  return(temp_K)
}
celsius_to_kelvin()

```

### 2. Why is there an error message when we ask R to print `vect1` to the console?
```{r, echo=TRUE, eval=FALSE}

celsius_to_kelvin <-function(temp_C) {
  temp_K <- temp_C + 273.15
  vect1 <- 1
  return(temp_K)
}

vect1 

```

# 5. Working with the tidyverse {.unnumbered}

Installing packages:
```{r, eval=FALSE}

#install.packages(“tidyverse”)
library(tidyverse)

```

## tidyverse grammar {.unnumbered}
```{r, eval=FALSE}

beaver2 %>%
  filter(activ == 1) %>%
  summarize(mean_temp = mean(temp))

```

## ggplot2 graphing {.unnumbered}

```{r, echo=TRUE}

library(ggplot2)
ggplot(beaver2,  
       aes(x = time, y = temp, color = activ)) + 
  geom_point()

```
