---
title: "Introduction to R workshop"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# 📘 1. Getting Started

R is a programming language -- in this workshop we'll use RStudio as an interface to explore data, write some code, and visualize results!

You can also use the cloud version for free - https://posit.cloud/plans/free. 

## Install R and RStudio

First, download both R (cran.r-project.org) and RStudio (posit.co/downloads/). 
Go to https://rstudio-education.github.io/hopr/starting.html for more information. 

# 🧩 2. Creating Variables

We use the `<-` operator to assign values to variables in R. 

## ✏️

### 1. Create a variable `my_age` in the code chunk below
```{r}

my_age <- 

```

### 2. Create a numeric vector `my_numbers` with three numbers
```{r}

my_numbers <- c(_, _, _)

```

### 3. 

## Dataset 

Take a look at the documentation on the ToothGrowth dataset in base R - note what the columns are named, the variables included, and the overall design of the study.

```{r, eval = FALSE}
?ToothGrowth
```

You can use the head() function to see the first 6 rows of the dataset. Each row in the dataset is an individual guinea pig, and the columns correspond to the response variable ('len') or the treatments ('supp' or 'dose').

```{r}
head(ToothGrowth)
```

# Base R {.unnumbered}

## Creating variables {.unnumbered}

In R, we create variables using the `<-` operator, which shows the direction in which the line of code is evaluated.

```{r}
weight_kg <- 55
myname <- "Borland"
```

Variables can be different types of objects - numbers, words, phrases. To check what a variable is classified as, we can call the `class()` function.

```{r}
class(weight_kg)
```

Now try the next few examples. When you're ready, change  `eval = FALSE` to `eval = TRUE`.

```{r, eval = FALSE}

my_age <- #your age
my_numbers <- c( , , ) #three numbers

example_var <- "55"
class(example_var)

```

## Exploring the dataset {.unnumbered}

Now back to our ToothGrowth dataset:

```{r}
class(ToothGrowth)
```

We can get some basic information about the dataset with a few simple functions:

```{r}
str(ToothGrowth)
summary(ToothGrowth)
```

## Subset the dataframe {.unnumbered}

We know that there are different treatment types, based on both supplement and dosage level of vitamin C, so it would make sense to evaluate the dataset based on these treatments. Thus, figuring out how to subset our dataframe is useful here.

We can subset our data by indexing:

```{r, eval=FALSE}

ToothGrowth[1, ] #pulls out the first row and all columns of ToothGrowth
ToothGrowth[ , 1] #all rows, but only the first column

```

Or by using logical conditions:

```{r, eval=FALSE}

ToothGrowth[ToothGrowth$dose > .5, ] #pulls out rows where the dosage 
#is greater than .5 mg/day, keeping all columns.

```

The `$` operator can be used to specify the name of a column in a dataframe when subsetting.

Now, try subsetting the dataset where the dosage is equal to 2 mg/day. Assign that answer to a new variable named `high_dose`.

## Base R graphing {.unnumbered}

Visualizing our data can help us understand the structure even further. 

### Simple graphs {.unnumbered}
```{r}
hist(ToothGrowth$len)

```

Now try using the `boxplot()` function to create a basic boxplot of our data. 

```{r}


```


### Less simple graph {.unnumbered}
Again, we probably want to look at our response variable (tooth length, or `len`) based on the different treatment types. We can make a boxplot to look at the spread.

```{r}
boxplot(len ~ supp * dose, data = ToothGrowth, ylab = "Length", 
        xlab = "Treatment and Dose")
# note the optional axis label arguments

```


## Basic statistics {.unnumbered}

```{r}
#creating a two-way ANOVA for our data
tg_aov <- aov(len ~ supp * dose, data = ToothGrowth)
summary(tg_aov)
```

### A note on assumptions

An ANOVA assumes: 1) response variable (`len` in our case) is quantitative and continuous; 2) variance is similar across groups; 3) observations are independent; and 4) data within groups is normally distributed. 

Conveniently, we can check two of these assumptions in R.

```{r}
#check assumption #2, variance 
# the plot() function is handy here - 
#the first output of our aov object will be the 
#residual vs. fitted graph
plot(tg_aov, 1)
#we want a horizontal red line

#check assumption #4, normality
# the second plot is the normal q-q plot
plot(tg_aov, 2)
#we want the points close to the diagonal line

```


# Writing functions {.unnumbered}.

# Working with the tidyverse {.unnumbered}

Installing packages:
```{r, eval=FALSE}

install.packages(“tidyverse”)
library(tidyverse)

```


## tidyverse grammar {.unnumbered}
```{r, eval=FALSE}
ToothGrowth %>%
  filter(dose == 2) %>%
  select(len)

```


## Manipulating data {.unnumbered}

## ggplot2 graphing {.unnumbered}

You can also embed plots, for example:

```{r}


```
